Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-05-18 manuscript/manuscript.Rmd
Now rendering figure captions

---
title: "Limits to Ecological Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning"
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  1. In the current age of a rapidly changing environment, it is becoming increasingly important to study critical transitions and how to best anticipate them. Critical transitions pose extremely challenging forecasting problems, which necessitate informative uncertainty estimation rather than point forecasts. In this study, we apply some of the most cutting edge deep learning methods for probabilistic time series forecasting to several classic ecological models that examine critical transitions.
  2. Our analysis focuses on three different simulated examples of critical transitions: a Hopf bifurcation, a saddle-node bifurcation and a stochastic transition. For each scenario, we compare the forecasts from four deep learning models, Long-short Term Memory networks, Gated Recurrent Unit networks, Block Recurrent neural networks and Transformers, to forecasts from an ARIMA model and a MCMC estimated model that is given the true transition dynamics.
  3. We found that the deep learning models were able to perform comparably to the idealized MCMC model on the stochastic transition case, and generally in between the MCMC and ARIMA models on the Hopf and saddle-node bifurcation examples.
  4. Our results establish that deep learning methods warrant further exploration on the challenging class of critical transition forecasting problems.

@@

```{r }
#| fig3
#| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
#| The bifurcation occurs soon after forecasting period begins, 
#| leading to progressively larger and larger oscillations. 
#| Prior to the bifurcation, pseudo-cycles are visible in the training data due to stochastic excitations. 
#| Following the bifurcation, stochasticity blurs the oscillatory pattern across replicate simulations. 
#| Only last 25 time points of training data are shown."

@@

```{r}
#| fig6
#| fig.cap="CRPS scores over time for each scenario.
#| Each line represents the scores against a replicate of time series observations from 'true model'. 
#| Patterns show forecast skill generally get worse over time - 
#| gradually in the case of Hopf bifurcation or suddenly in repsonse to the saddle-node bifurcation. 
#| The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
#| In the stochastic transition context, this creates two branches, where high values indicate periods of 
#| time when the forecast predicts the wrong state and lower branch indicates the correct state."

>>>>

---
title: "Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning"
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  1. In the current age of a rapidly changing environment, it is becoming increasingly important to study critical transitions and how to best anticipate them. Critical transitions pose extremely challenging forecasting problems, which necessitate informative uncertainty estimation rather than point forecasts. In this study, we apply some of the most cutting edge deep learning methods for probabilistic time series forecasting to several classic ecological models that examine critical transitions.
  2. Our analysis focuses on three different simulated examples of critical transitions: a Hopf bifurcation, a saddle-node bifurcation and a stochastic transition. For each scenario, we compare the forecasts from four deep learning models, Long-short Term Memory networks, Gated Recurrent Unit networks, Block Recurrent neural networks and Transformers, to forecasts from an ARIMA model and a MCMC estimated model that is given the true transition dynamics.
  3. We found that the deep learning models were able to perform comparably to the idealized MCMC model on the stochastic transition case, and generally in between the MCMC and ARIMA models on the Hopf and saddle-node bifurcation examples.
  4. Our results establish that deep learning methods warrant further exploration on the challenging class of critical transition forecasting problems.

@@

```{r}
#| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
#| The bifurcation occurs soon after forecasting period begins, 
#| leading to progressively larger and larger oscillations. 
#| Prior to the bifurcation, pseudo-cycles are visible in the training data due to stochastic excitations. 
#| Following the bifurcation, stochasticity blurs the oscillatory pattern across replicate simulations. 
#| Only last 25 time points of training data are shown."

@@

```{r}
#| fig.cap="CRPS scores over time for each scenario.
#| Each line represents the scores against a replicate of time series observations from 'true model'. 
#| Patterns show forecast skill generally get worse over time - 
#| gradually in the case of Hopf bifurcation or suddenly in repsonse to the saddle-node bifurcation. 
#| The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
#| In the stochastic transition context, this creates two branches, where high values indicate periods of 
#| time when the forecast predicts the wrong state and lower branch indicates the correct state."
